---
title: "survey data import and harmonization for mid-Atlantic process models"
author: "Alexa Fredston-Hermann"
date: "5/7/2020"
output: html_document
---

This script is based on previous code M. Pinsky wrote to pre-process data for the Atlantic cod forecast challenge, and on compile.R in the OceanAdapt data files. We first crop the trawl data down to the regions and species of interest, and then revisit the raw data files to extract length frequency data. 

Users should specify:

1. The file path to a directory containing a download of OceanAdapt, `OApath`
1. A list of Latin names (format Genus species) of species of interest
1. A list of regions of interest, from: "Aleutian Islands", "Eastern Bering Sea", "Gulf of Alaska", 
"Northeast US Spring", "Northeast US Fall", "West Coast Triennial", 
"West Coast Annual", "Gulf of Mexico", "Southeast US Spring", 
"Southeast US Summer", "Southeast US Fall", "Scotian Shelf"

```{r packages and functions, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
library(PBSmapping)
library(tidyverse)
library(here)
library(stringr)
library(lubridate)
library(data.table)
here <- here::here

OApath <- "~/github/OceanAdapt_9815545/"
# sppOfInt <- c("Squalus acanthias")
regOfInt <- c("Northeast US Spring", "Northeast US Fall", "Southeast US Spring", "Southeast US Summer", "Southeast US Fall", "Scotian Shelf")
# options: c("Aleutian Islands", "Eastern Bering Sea", "Gulf of Alaska", "Northeast US Spring", "Northeast US Fall", "West Coast Triennial", "West Coast Annual", "Gulf of Mexico", "Southeast US Spring", "Southeast US Summer", "Southeast US Fall", "Scotian Shelf")

```


# Get zero-inflated data from OceanAdapt, and other datasets
```{r data, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
load(paste0(OApath,"data_clean/dat_exploded.RData"))

```

# Get length-frequency data from individual surveys

## Northeast US

```{r neus len, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
load(paste0(OApath,"data_raw/neus_Survdat.RData"))
load(paste0(OApath, "data_raw/neus_SVSPP.RData"))

dat_exploded_neus <- dat.exploded %>% 
  filter(region %in% c("Northeast US Spring", "Northeast US Fall"))

neus_len <- survdat %>% 
  # create a haulid for joining with dat.exploded
  mutate(haulid = paste(formatC(CRUISE6, width=6, flag=0), formatC(STATION, width=3, flag=0), formatC(STRATUM, width=4, flag=0), sep='-')) %>% 
  select(haulid, SVSPP, LENGTH, NUMLEN) %>% 
  filter(haulid %in% dat_exploded_neus$haulid) %>% # pare down to hauls of interest 
  left_join(spp, by="SVSPP") %>% # get species names from species codes
  mutate(spp = str_to_sentence(SCINAME)) %>% # change to format of sppOfInt
  select(spp, haulid, LENGTH, NUMLEN) %>%
  filter(!is.na(LENGTH))

# creating a separate bottom temperature dataframe so it can be joined at the end after rdoing all the length class sorting
btemp <- survdat %>% 
  # create a haulid for joining with dat.exploded
  mutate(haulid = paste(formatC(CRUISE6, width=6, flag=0), formatC(STATION, width=3, flag=0), formatC(STRATUM, width=4, flag=0), sep='-')) %>% 
  select(haulid, BOTTEMP) %>% 
  distinct() %>%
  rename("btemp"=BOTTEMP)
  
```

## Scotian Shelf

Doesn't look like this has any length data--just total number and total weight. 

```{r scot len, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
scotfiles <- as.list(dir(pattern = "scot", path = "~/github/OceanAdapt_9815545/data_raw", full.names = T))

scotraw <- scotfiles %>% 
  map_dfr(~ read_csv(.x, col_types = cols(
    .default = col_double(),
    MISSION = col_character(),
    SEASON = col_character(),
    SURVEYDATE = col_character(),
    GEAR = col_character(),
    SCIENTIFICNAME = col_character(),
    TAXONOMICNAMEAUTHOR = col_character()
  ))) 
```

## Southeast US reef fish survey

```{r serfs, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}

#SERFS
serfs_len_raw <- read.delim(here::here("survey-data","southeast_reef_fish_survey_length_frequency_04_10_2020_raw.txt"), sep=",")

serfs_len_tidy <- serfs_len_raw %>%
  rename_all(tolower) %>% 
  mutate_all(tolower) %>% # so dataframe isn't screaming at me
  mutate_all(funs(str_replace(., "=", ""))) %>% # raw data has random equals signs everywhere
  mutate_all(na_if, "not applicable") %>% 
  select_if(~!all(is.na(.))) %>%  # remove columns that just have NAs in them 
  distinct() %>% 
  mutate(date = mdy(date)) %>% 
  select(eventname, date, speciesscientificname, speciescommonname, length, numbermeasure, totalnumber, latitudestart, longitudestart) %>% 
  distinct() %>% 
  rename("sppCommon"=speciescommonname,
         "sppLatin"=speciesscientificname) %>% 
  mutate(length = as.numeric(length),
         totalnumber = as.numeric(totalnumber)) %>% 
  filter(!is.na(length), !is.na(totalnumber)) %>% # just removing one weird NA row
  mutate(year = lubridate::year(date),
         haulid = eventname) %>% 
  select(haulid, year, sppLatin, length, numbermeasure, totalnumber, latitudestart, longitudestart)
```

## Southeast US trawl

This survey has a lot of weight-related measurements and I'm not totally sure how they relate to each other: SPECIESTOTALWEIGHT, SPECIESSUBWEIGHT, SPECIESWGTPROCESSED, CATCHSUBSAMPLED, CATCHWEIGHT, CATCHSUBWEIGHT. 

```{r seus oa, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}

seus_catch <- read_csv(unz("~/github/OceanAdapt_9815545/data_raw/seus_catch.csv.zip", "seus_catch.csv"), col_types = cols(.default = col_character())) %>% 
  # remove symbols
  mutate_all(list(~str_replace(., "=", ""))) %>% 
  mutate_all(list(~str_replace(., '"', ''))) %>% 
  mutate_all(list(~str_replace(., '"', ''))) %>%
  mutate(spp = str_to_sentence(SPECIESSCIENTIFICNAME))
```

As per Zoe Kitchel and Ria Kobernuss, the [SEAMAP online data portal](https://www2.dnr.sc.gov/seamap/Account/LogOn?ReturnUrl=%2fseamap%2fReports) does have an option to download finer-scale data that includes length data. They shared it with me (May 14 2020) with the following note, and it's imported below:

*This gives you a CSV. I converted it to an RData file last September, which goes up until 11/16/2018. I just checked, and it hasn't been updated since then, so the RData file we've been using is still up to date.*

```{r seus manual, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE }
load(here("survey-data","seus_trawl_data.Rdata"))

seus_len <- seus_specimens %>%
  rename_all(tolower) %>% 
  mutate_all(tolower) %>% # so dataframe isn't screaming at me
  mutate_all(funs(str_replace(., "=", ""))) %>% # raw data has random equals signs everywhere
  mutate_all(funs(str_replace(., '"', ''))) %>% # also get rid of quotation marks
  mutate_all(funs(str_replace(., '"', ''))) %>% # have to run twice to get rid of leading and trailing quotes, for some reason 
  # mutate_all(na_if, "not applicable") %>% 
  select_if(~!all(is.na(.))) # remove columns that are all NAs

```

I'm not sure how to use the Southeast data because (unlike Northeast) length measurements don't appear to have been recorded on *every* haul, so we won't always have associated length data for positive catches. They're also only measured for a subset of species that (of our study species) only includes spiny dogfish.

# Summarize length data into length classes CURRENTLY NEUS ONLY BUT BOTH SEASONS

Note that the chunk below is written for one species only (spiny dogfish)--will need to be updated in the future when more species are added in. 

```{r dogfish,  eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
dogfish_neus_len <- neus_len %>% filter(spp=="Squalus acanthias") 

dogfish_lm_F <- 75 #cm; see SAW 43 p17
dogfish_lm_M <- 60 

dogfish_lm <- mean(c(dogfish_lm_M, dogfish_lm_F)) # 67.5
dogfish_ljuv <- min(dogfish_neus_len$LENGTH) + (dogfish_lm - min(dogfish_neus_len$LENGTH))/2 # halfway between min size caught and Lm; 39.75

dogfish_neus_len_agg <- dogfish_neus_len %>%
  mutate(lengthclass = ifelse(LENGTH <= dogfish_ljuv, "smalljuv",
                              ifelse(LENGTH < dogfish_lm, "largejuv", "adult"))) %>% # group lengths into length classes 
  group_by(haulid, spp, lengthclass) %>% 
  summarise(numlengthclass = sum(NUMLEN)) # aggregate counts

dogfish_neus_len_zeros <- expand_grid(haulid=unique(dat_exploded_neus$haulid),  lengthclass=unique(dogfish_neus_len_agg$lengthclass),
                                 spp="Squalus acanthias") %>% # get all possible combinations of haulid and length class 
  left_join(dogfish_neus_len_agg, by=c("spp","haulid","lengthclass")) %>% # add length counts 
  left_join(dat_exploded_neus, by=c("spp","haulid")) %>% # add haul data 
  left_join(btemp, by="haulid") %>%
  select(-wtcpue) %>% # not accurate after all the joins/expansions
  mutate(numlengthclass = replace_na(numlengthclass, 0)) # replace NAs with true zeroes
```

Generating the same datasets for the other three focal species, in order to visualize their data distributions. So far just using NEUS for summer flounder (which is also found in the SEAMAP survey) and shortfin squid (which I think is actually constrained just to the NEUS survey).

**Revisit break points for size classes and data inputs for these three!**

## Summer flounder

I'm adding in the survey selectivity correction before transforming lengths to classes, because that's how Mark communicated it to me:

The current assessment model provides estimates of NEFSC HB Bigelow survey selectivity at age for 2009-2019.
For spring: age 0 = 0.00, age 1 = 0.75, age 2 = 0.80, age 3-7+ = 1.00 
For fall:      age 0 = 0.72, age 1 = 0.80, age 2 = 1.00, age 3-7+ = 1.00

As a rough estimation of how ages relate to lengths:
In the spring no age 0 fish are caught, age 1 fish are generally < 36 cm, age 2 fish are 36 - 40 cm, and age 3 fish are > 40 cm.
In the fall, age 0 fish are < 30 cm, age 1 fish are 30-40 cm, age 2 fish are 41-45 cm, and age 3 fish are > 45 cm.

```{r summer flounder,  eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
fluke_neus_len <- neus_len %>% filter(spp=="Paralichthys dentatus", LENGTH>5) # getting rid of  a single fish at 4cm that was throwing off the stage 1/2 breakpoint--after that there are a decent number of 9, 10, 11... cm fish

fluke_lm <- 27 # p64 SAW 66
fluke_ljuv <- min(fluke_neus_len$LENGTH) + (fluke_lm - min(fluke_neus_len$LENGTH))/2

# incorporating size selectivity
fluke_neus_fall <- fluke_neus_len %>% 
  left_join(dat_exploded_neus %>% select(region, haulid) %>% distinct(), by="haulid") %>% #to get seasons
  filter(region=="Northeast US Spring") %>% 
  mutate(numlen_correct = ifelse(LENGTH<36, NUMLEN*(1/0.75), ifelse(LENGTH<40, NUMLEN*(1/0.80), NUMLEN)))

fluke_neus_spring <- fluke_neus_len %>% 
  left_join(dat_exploded_neus %>% select(region, haulid) %>% distinct(), by="haulid") %>% #to get seasons
  filter(region=="Northeast US Spring") %>% 
  mutate(numlen_correct = ifelse(LENGTH<36, NUMLEN*(1/0.75), ifelse(LENGTH<40, NUMLEN*(1/0.80), NUMLEN)))

fluke_neus_len_agg <- fluke_neus_len %>%
  mutate(lengthclass = ifelse(LENGTH <= fluke_ljuv, "smalljuv",
                              ifelse(LENGTH < fluke_lm, "largejuv", "adult"))) %>% 
  group_by(haulid, spp, lengthclass) %>% 
  summarise(numlengthclass = sum(NUMLEN))  

fluke_neus_len_zeros <- expand_grid(haulid=unique(dat_exploded_neus$haulid),  lengthclass=unique(fluke_neus_len_agg$lengthclass),
                                 spp="Paralichthys dentatus") %>%  
  left_join(fluke_neus_len_agg, by=c("spp","haulid","lengthclass")) %>% 
  left_join(dat_exploded_neus, by=c("spp","haulid")) %>% 
  left_join(btemp, by="haulid") %>%
  select(-wtcpue) %>% 
  mutate(numlengthclass = replace_na(numlengthclass, 0)) 
```

```{r shortfin squid,  eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
squid_neus_len <- neus_len %>% filter(spp=="Illex illecebrosus")

squid_lm <- 20 # FOR EXPLORATORY PURPOSES ONLY! based on a best guess from this sentence in the FAO page: "Summer hatchlings achieve a mantle length of about 18 or 19 cm after 1 year (females are slightly larger than males) and are ready to spawn in winter at an age of about 1 ½ years (Mesnil, 1977)." http://www.fao.org/fishery/species/2720/en
squid_ljuv <- min(squid_neus_len$LENGTH) + (squid_lm - min(squid_neus_len$LENGTH))/2

squid_neus_len_agg <- squid_neus_len %>%
  mutate(lengthclass = ifelse(LENGTH <= squid_ljuv, "smalljuv",
                              ifelse(LENGTH < squid_lm, "largejuv", "adult"))) %>% 
  group_by(haulid, spp, lengthclass) %>% 
  summarise(numlengthclass = sum(NUMLEN))  

squid_neus_len_zeros <- expand_grid(haulid=unique(dat_exploded_neus$haulid),  lengthclass=unique(squid_neus_len_agg$lengthclass),
                                 spp="Illex illecebrosus") %>%  
  left_join(squid_neus_len_agg, by=c("spp","haulid","lengthclass")) %>% 
  left_join(dat_exploded_neus, by=c("spp","haulid")) %>% 
  left_join(btemp, by="haulid") %>%
  select(-wtcpue) %>% 
  mutate(numlengthclass = replace_na(numlengthclass, 0)) 
```

```{r gray triggerfish,  eval=TRUE, echo=FALSE, results=FALSE, message=FALSE}
# SERFS 
gray_serfs_len <- serfs_len_tidy %>% 
  filter(sppLatin=="balistes capriscus")

# using SEDAR reported Lm values, and converting from FL to TL *in millimeters*, then converting to cm 
gray_t_lm_TL_female <- (-18.27+1.21*177)/10
gray_t_lm_TL_male <- (-18.27+1.21*180)/10
gray_lm <- mean(c(gray_t_lm_TL_female,gray_t_lm_TL_male))
gray_ljuv <- min(gray_serfs_len$length) + (gray_lm - min(gray_serfs_len$length))/2 # need to think more about how to define this based on multiple surveys. minimum across all of them? in SERFS it's 10 

gray_serfs_len_agg <- gray_serfs_len %>%
  mutate(lengthclass = ifelse(length <= gray_ljuv, "smalljuv",
                              ifelse(length < gray_lm, "largejuv", "adult"))) %>% 
  group_by(haulid, sppLatin, lengthclass) %>% 
  summarise(numlengthclass = sum(totalnumber))  

gray_serfs_len_zeros <- expand_grid(haulid=unique(serfs_len_tidy$haulid),  lengthclass=unique(gray_serfs_len_agg$lengthclass),
                                 sppLatin="balistes capriscus") %>%  
  left_join(gray_serfs_len_agg, by=c("sppLatin","haulid","lengthclass")) %>% 
left_join(serfs_len_tidy %>% select(haulid, year, latitudestart, longitudestart) %>% distinct(), by="haulid") %>% # adding haul info back in 
  mutate(numlengthclass = replace_na(numlengthclass, 0)) 

# since this species isn't in the SEUS length-frequency data, let's skip it for now

#NEUS
# gray_neus_len <- neus_len %>% filter(spp=="Balistes capriscus") # 751 observations, not that helpful... min len 1??
# 
# gray_neus_len_agg <- gray_neus_len %>%
#   mutate(lengthclass = ifelse(LENGTH <= gray_ljuv, "smalljuv",
#                               ifelse(LENGTH < gray_lm, "largejuv", "adult"))) %>% 
#   group_by(haulid, spp, lengthclass) %>% 
#   summarise(numlengthclass = sum(NUMLEN))  
# 
# # this is not joining properly for some reason
# gray_neus_len_zeros <- expand_grid(haulid=unique(dat_exploded_neus$haulid),  lengthclass=unique(gray_neus_len_agg$lengthclass),
#                                  spp="Balistes capriscus") %>%  
#   left_join(gray_neus_len_agg, by=c("spp","haulid","lengthclass")) %>% 
#   left_join(dat_exploded_neus, by=c("spp","haulid")) %>% 
#    left_join(btemp, by="haulid") %>% 
#   select(-wtcpue) %>% 
#   mutate(numlengthclass = replace_na(numlengthclass, 0)) 

```


# Write out data

In a format that can be imported into Python

```{r export df, eval=TRUE, echo=FALSE, results=FALSE, message=FALSE }
# this csv is too big for github, so zip it
write_csv(dogfish_neus_len_zeros, here::here("processed-data", "dogfish_prepped_data.csv"))
write_csv(fluke_neus_len_zeros, here::here("processed-data", "fluke_prepped_data.csv"))
write_csv(squid_neus_len_zeros, here::here("processed-data", "squid_prepped_data.csv"))
write_csv(gray_serfs_len_zeros, here::here("processed-data", "gray_prepped_data.csv"))

zip(zipfile = here::here("processed-data", "dogfish_prepped_data.zip"),
    files = here::here("processed-data", "dogfish_prepped_data.csv"))
```

